---
title: "Credit Card Fraud Detection"
author: "Suin Kim"
output: html_document
---

# 1. Setup & Data Loading

```{r setup, message=FALSE, warning=FALSE}
# Load all required libraries upfront
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(stats)
library(pROC)
library(randomForest)
library(ranger)
library(caret)

# Load dataset
# Place card_transdata.csv in your working directory before running
# Dataset: https://www.kaggle.com/datasets/dhanushnarayananr/credit-card-fraud
df <- read_csv("card_transdata.csv", show_col_types = FALSE)

cat("Dimensions:", dim(df), "\n")
```

---

# 2. Exploratory Data Analysis

## 2.1 Dataset Overview

```{r eda-overview}
# Column summary: data types, missing values, and averages
summary_table <- df %>%
  summarise(
    Column_Name = names(.),
    Data_Type = sapply(., class),
    Missing_Values = sapply(., function(col) sum(is.na(col))),
    Average = sapply(., function(col) if (is.numeric(col)) round(mean(col, na.rm = TRUE), 4) else NA)
  ) %>%
  as.data.frame()

kable(summary_table, caption = "Column Summary: Data Type, Missing Values, and Averages")
```

```{r eda-summary}
# Full statistical summary of all variables
summary(df)
```

## 2.2 Class Distribution

```{r eda-class-distribution}
# The target variable (fraud) is heavily imbalanced at ~8.74% fraud rate.
# This needs to be addressed before modeling to avoid biased predictions.
fraud_counts <- table(df$fraud)

barplot(fraud_counts,
        main = "Fraud Count Distribution",
        xlab = "Fraud Indicator",
        ylab = "Number of Transactions",
        col = c("green", "red"),
        names.arg = c("No Fraud", "Fraud"))

cat("Fraud rate:", round(mean(df$fraud) * 100, 2), "%\n")
```

## 2.3 Fraud Rate by Binary Variables

```{r eda-fraud-rates}
# Before modeling, we examine fraud rates across binary features.
# This helps us understand which variables carry the most predictive signal.
binary_vars <- c("repeat_retailer", "used_chip", "used_pin_number", "online_order")

fraud_rate_table <- do.call(rbind, lapply(binary_vars, function(var) {
  df %>%
    group_by(Value = .data[[var]]) %>%
    summarise(Fraud_Rate = round(mean(fraud) * 100, 2), .groups = "drop") %>%
    mutate(Variable = var) %>%
    select(Variable, Value, Fraud_Rate)
}))

kable(fraud_rate_table, caption = "Fraud Rate by Binary Variable")
```

Key findings:
- **used_pin_number** is the strongest signal — fraud rate drops from 9.7% to just 0.27% when PIN is used
- **online_order** nearly 10x the fraud rate compared to in-person transactions (12.7% vs 1.3%)
- **repeat_retailer** shows almost no difference — motivating the interaction term tested later

## 2.4 Continuous Variable Distributions

```{r eda-histograms-raw}
# Raw distributions — all three variables are heavily right-skewed
# with extreme outliers, making log transformation necessary
variables <- c("distance_from_home", "distance_from_last_transaction", "ratio_to_median_purchase_price")

par(mfrow = c(1, 3), mar = c(5, 4, 4, 2))
for (var in variables) {
  hist(df[[var]],
       main = paste("Raw:", var),
       xlab = var,
       col = "grey",
       border = "black",
       breaks = 25)
}
par(mfrow = c(1, 1))
```

```{r eda-histograms-transformed}
# After log + sqrt transformation, distributions are more symmetric
# and suitable for modeling
par(mfrow = c(1, 3), mar = c(5, 4, 4, 2))
for (var in variables) {
  log_var <- (log(df[[var]] + 1))^(0.5)
  hist(log_var,
       main = paste("Transformed:", var),
       xlab = var,
       col = "grey",
       border = "black",
       breaks = 25)
}
par(mfrow = c(1, 1))
```

## 2.5 Fraud vs Non-Fraud: Continuous Variables

```{r eda-boxplots}
# Box plots comparing fraud vs non-fraud across continuous variables.
# Clear separation confirms these variables are meaningful predictors.
aliases <- c(
  "distance_from_home" = "Distance from Home",
  "distance_from_last_transaction" = "Distance from Last Transaction",
  "ratio_to_median_purchase_price" = "Ratio to Median Purchase Price"
)

for (var in variables) {
  p <- ggplot(df, aes(x = factor(fraud, levels = c(0, 1), labels = c("No", "Yes")),
                      y = log(df[[var]] + 1)^(0.5), fill = factor(fraud))) +
    geom_boxplot(outlier.shape = NA) +
    labs(
      title = paste("Fraud Status Comparison:", aliases[var]),
      x = "Fraud Status",
      y = paste("Transformed", aliases[var])
    ) +
    scale_fill_manual(values = c("green", "red"), name = "Fraud Status") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  print(p)
}
```

## 2.6 Binary Variables vs Fraud

```{r eda-mosaic}
# Mosaic plots show the proportion of fraud within each binary category.
# Larger red areas indicate higher fraud concentration.
aliases_binary <- c(
  "repeat_retailer" = "Repeat Retailer",
  "used_chip" = "Used Chip",
  "used_pin_number" = "Used PIN",
  "online_order" = "Online Order"
)

par(mfrow = c(2, 2), mar = c(5, 4, 4, 2))
for (var in binary_vars) {
  contingency_table <- table(df[[var]], df$fraud)
  mosaicplot(
    contingency_table,
    main = paste("Mosaic:", aliases_binary[var], "vs Fraud"),
    xlab = aliases_binary[var],
    ylab = "Fraud",
    color = TRUE
  )
}
par(mfrow = c(1, 1))
```

## 2.7 Correlation Matrix

```{r eda-correlation}
# Check for multicollinearity between predictors.
# High correlation between distance_from_home and repeat_retailer
# motivates testing an interaction term in the logistic regression.
numeric_cols <- sapply(df, is.numeric)
cor_matrix <- cor(df[, numeric_cols], use = "complete.obs")
print(round(cor_matrix, 3))
```

```{r eda-interaction-check}
# Visualizing distance_from_home vs repeat_retailer to assess
# whether an interaction term is warranted in the model
p <- ggplot(df, aes(x = factor(repeat_retailer, levels = c(0, 1), labels = c("No", "Yes")),
                    y = log(df$distance_from_home)^(0.5), fill = factor(repeat_retailer))) +
  geom_boxplot(outlier.shape = NA) +
  labs(
    title = "Distance from Home vs Repeat Retailer",
    x = "Repeat Retailer",
    y = "Transformed Distance from Home"
  ) +
  scale_fill_manual(values = c("lightblue", "lightgreen"), name = "Repeat Retailer") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
print(p)
```

---

# 3. Data Cleaning & Balancing

```{r balancing}
# The dataset is imbalanced (8.74% fraud). Training on imbalanced data causes
# the model to predict the majority class almost exclusively.
# We use bootstrapping (oversampling with replacement) to balance the classes.

# Check original distribution
cat("Original class distribution:\n")
print(table(df$fraud))

# Separate minority and majority classes
minority <- df[df$fraud == 1, ]
majority <- df[df$fraud == 0, ]

# Bootstrap minority class to match majority size
set.seed(123)
minority_boot <- minority[sample(1:nrow(minority), size = nrow(majority), replace = TRUE), ]

# Combine into balanced dataset
balanced_df <- rbind(minority_boot, majority)

cat("\nBalanced class distribution:\n")
print(table(balanced_df$fraud))
```

```{r balancing-plot}
# Confirm balance visually
fraud_counts_balanced <- table(balanced_df$fraud)
barplot(fraud_counts_balanced,
        main = "Fraud Distribution After Bootstrapping",
        xlab = "Fraud Indicator",
        ylab = "Number of Transactions",
        col = c("green", "red"),
        names.arg = c("No Fraud", "Fraud"))
```

---

# 4. Train / Test Split

```{r train-test-split}
# 70/30 train-test split on the balanced dataset
# set.seed ensures reproducibility
set.seed(123)
train_ratio <- 0.7
train_indices <- sample(1:nrow(balanced_df), size = floor(train_ratio * nrow(balanced_df)))

train_data <- balanced_df[train_indices, ]
test_data  <- balanced_df[-train_indices, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size: ", nrow(test_data), "\n")
cat("\nTraining class distribution:\n")
print(table(train_data$fraud))
cat("\nTesting class distribution:\n")
print(table(test_data$fraud))
```

---

# 5. Logistic Regression

## 5.1 Feature Selection via Backward Elimination

```{r logistic-backward}
# Start with all predictors and use AIC-based backward elimination
# to identify the most informative subset of features.
full_model <- glm(fraud ~ ., data = train_data, family = binomial)
best_model <- step(full_model, direction = "backward")
summary(best_model)
```

## 5.2 Base Model (No Interaction)

```{r logistic-base}
# Fit logistic regression with all predictors selected by backward elimination.
# This serves as the baseline before testing interaction terms.
model <- glm(fraud ~ distance_from_home + distance_from_last_transaction +
               ratio_to_median_purchase_price + repeat_retailer + used_chip +
               used_pin_number + online_order,
             data = train_data, family = binomial)

summary(model)

# Predict on training data
predicted_probs <- predict(model, type = "response")
predicted_classes <- ifelse(predicted_probs > 0.5, 1, 0)

cat("\nTraining Confusion Matrix:\n")
print(table(Predicted = predicted_classes, Actual = train_data$fraud))

# Fitted probability plot — well-separated distributions indicate good model fit
ggplot(train_data, aes(x = predicted_probs, fill = factor(fraud))) +
  geom_density(alpha = 0.5) +
  labs(title = "Fitted Probability Plot (Base Model)",
       x = "Predicted Probability",
       fill = "Fraud (0 = No, 1 = Yes)") +
  theme_minimal()
```

## 5.3 Interaction Model

```{r logistic-interaction}
# EDA showed correlation between distance_from_home and repeat_retailer.
# We test whether their interaction improves model fit.
model2 <- glm(fraud ~ distance_from_home + distance_from_last_transaction +
                ratio_to_median_purchase_price + repeat_retailer + used_chip +
                used_pin_number + online_order + (distance_from_home * repeat_retailer),
              data = train_data, family = binomial)

summary(model2)

predicted_probs_m2 <- predict(model2, type = "response")
predicted_classes_m2 <- ifelse(predicted_probs_m2 > 0.5, 1, 0)

cat("\nTraining Confusion Matrix (Interaction Model):\n")
print(table(Predicted = predicted_classes_m2, Actual = train_data$fraud))

ggplot(train_data, aes(x = predicted_probs_m2, fill = factor(fraud))) +
  geom_density(alpha = 0.5) +
  labs(title = "Fitted Probability Plot (Interaction Model)",
       x = "Predicted Probability",
       fill = "Fraud (0 = No, 1 = Yes)") +
  theme_minimal()
```

## 5.4 Model Comparison: Likelihood Ratio Test

```{r logistic-lrt}
# Likelihood Ratio Test compares the two models.
# A significant p-value means the interaction term adds meaningful explanatory power.
model_no_interaction <- glm(fraud ~ distance_from_home + distance_from_last_transaction +
                               ratio_to_median_purchase_price + repeat_retailer + used_chip +
                               used_pin_number + online_order,
                             family = "binomial", data = balanced_df)

model_interaction <- glm(fraud ~ distance_from_home + distance_from_last_transaction +
                            ratio_to_median_purchase_price + repeat_retailer + used_chip +
                            used_pin_number + online_order + (distance_from_home * repeat_retailer),
                          family = "binomial", data = balanced_df)

anova(model_no_interaction, model_interaction, test = "LRT")
```

## 5.5 Test Set Evaluation

```{r logistic-test-eval}
# Evaluate the base model on held-out test data.
# Precision and recall are more informative than accuracy for imbalanced fraud data.
test_probs <- predict(model, newdata = test_data, type = "response")
test_preds <- ifelse(test_probs > 0.5, 1, 0)

conf_matrix <- table(Predicted = test_preds, Actual = test_data$fraud)
print(conf_matrix)

accuracy  <- sum(diag(conf_matrix)) / sum(conf_matrix)
precision <- conf_matrix[2, 2] / sum(conf_matrix[2, ])
recall    <- conf_matrix[2, 2] / sum(conf_matrix[, 2])
f1_score  <- 2 * (precision * recall) / (precision + recall)

cat("Accuracy: ", round(accuracy, 4), "\n")
cat("Precision:", round(precision, 4), "\n")
cat("Recall:   ", round(recall, 4), "\n")
cat("F1-Score: ", round(f1_score, 4), "\n")
```

```{r logistic-roc}
# ROC curve and AUC measure discrimination ability across all thresholds.
# AUC of 1.0 = perfect, 0.5 = random chance.
roc_obj <- roc(test_data$fraud, test_preds)
auc_val <- auc(roc_obj)
plot(roc_obj, main = paste("ROC Curve (AUC =", round(auc_val, 3), ")"))
cat("AUC:", round(auc_val, 3), "\n")
```

---

# 6. Random Forest

```{r rf-model}
# Random Forest is an ensemble method that builds many decision trees and
# aggregates their predictions. It handles non-linear relationships and
# interactions naturally, making it well-suited for fraud detection.

# Convert fraud to factor for classification
df$fraud <- as.factor(df$fraud)

# Train/test split on original (unbalanced) df for RF
# RF handles class imbalance better than logistic regression
set.seed(123)
train_index_rf <- createDataPartition(df$fraud, p = 0.7, list = FALSE)
train_data_rf  <- df[train_index_rf, ]
test_data_rf   <- df[-train_index_rf, ]

# Fit RF with ranger (memory-efficient implementation)
# num.trees=200 balances accuracy and compute time
# mtry=3 = number of variables tried at each split
rf_model <- ranger(fraud ~ ., data = train_data_rf,
                   num.trees = 200, mtry = 3, importance = "impurity")

print(rf_model)

# Predict on test data
rf_predictions      <- predict(rf_model, data = test_data_rf, type = "response")
rf_predicted_classes <- rf_predictions$predictions

# Confusion matrix
confusionMatrix(as.factor(rf_predicted_classes), test_data_rf$fraud)
# Note: Near-perfect accuracy is expected here but warrants cross-validation
# to rule out overfitting
```

```{r rf-importance}
# Feature importance shows which variables drive predictions the most.
# Higher impurity reduction = more important variable.
importance <- rf_model$variable.importance
sorted_importance <- sort(importance, decreasing = TRUE)

par(mar = c(5, 11, 8, 2))
barplot(
  sorted_importance,
  main = "Random Forest Feature Importance",
  horiz = TRUE,
  las = 1,
  col = "steelblue",
  xlab = "Importance Score",
  names.arg = names(sorted_importance),
  cex.names = 0.8
)
par(mar = c(5, 4, 4, 2))
```

## 6.1 Cross-Validation

```{r rf-cv}
# The initial RF hit near-100% accuracy — a red flag for overfitting.
# Cross-validation confirms whether this holds across different data splits.
cv_control <- trainControl(
  method = "cv",
  number = 3,
  verboseIter = TRUE,
  savePredictions = "final"
)

rf_cv_model <- train(
  fraud ~ ., data = train_data_rf,
  method = "ranger",
  trControl = cv_control,
  tuneGrid = expand.grid(
    mtry = c(2, 3),
    splitrule = "gini",
    min.node.size = c(5)
  ),
  metric = "Accuracy",
  num.trees = 200
)

print(rf_cv_model)

# Predict on test data using CV model
cv_test_preds <- predict(rf_cv_model, newdata = test_data_rf)
cv_conf_matrix <- confusionMatrix(cv_test_preds, test_data_rf$fraud)
print(cv_conf_matrix)
```

---

# 7. Example Predictions

```{r example-transactions}
# To demonstrate real-world utility, we test three transactions:
# a clearly safe one, a clearly fraudulent one, and a borderline case.
# This shows how the model behaves across different risk profiles.

example_transactions <- data.frame(
  scenario = c("Low Risk", "High Risk", "Borderline"),
  distance_from_home = c(2, 150, 25),
  distance_from_last_transaction = c(1, 200, 13),
  ratio_to_median_purchase_price = c(0.8, 8.5, 1.5),
  repeat_retailer = c(1, 0, 1),
  used_chip = c(1, 0, 1),
  used_pin_number = c(1, 0, 0),
  online_order = c(0, 1, 1)
)

# Predict fraud probability for each scenario using the interaction model
input_data <- example_transactions[, -1]  # drop scenario label column
predicted_probs <- predict(model2, newdata = input_data, type = "response")
predicted_classes <- ifelse(predicted_probs > 0.5, "Fraud", "Not Fraud")

results <- example_transactions %>%
  mutate(
    Fraud_Probability = round(predicted_probs, 4),
    Prediction = predicted_classes
  )

kable(results, caption = "Example Transaction Predictions")
```

**Interpretation:**
- **Low Risk**: Short distance from home, PIN used, chip used, in-person — model assigns near-zero fraud probability. Matches EDA finding that PIN usage drops fraud rate to 0.27%.
- **High Risk**: Far from home, far from last transaction, high purchase ratio, online, no chip, no PIN — model flags as high fraud probability. Every major risk factor is present.
- **Borderline**: Mixed signals — repeat retailer and chip used, but online order with no PIN and moderate distance. Model uses the interaction between `distance_from_home` and `repeat_retailer` to moderate the risk.

---

## 7.1 Threshold Analysis

```{r threshold-analysis}
# The default 0.5 threshold is not always optimal for fraud detection.
# Lowering the threshold catches more fraud (higher recall) but increases
# false positives — legitimate transactions flagged as fraud.
# This is a business decision: what is the cost of a missed fraud vs
# the cost of falsely declining a legitimate transaction?

thresholds <- c(0.3, 0.4, 0.5, 0.6, 0.7)

threshold_results <- do.call(rbind, lapply(thresholds, function(t) {
  preds <- ifelse(test_probs > t, 1, 0)
  cm <- table(Predicted = preds, Actual = test_data$fraud)
  
  # Handle edge cases where a class may be missing
  tp <- ifelse(!is.na(cm["1","1"]), cm["1","1"], 0)
  fp <- ifelse(!is.na(cm["1","0"]), cm["1","0"], 0)
  tn <- ifelse(!is.na(cm["0","0"]), cm["0","0"], 0)
  fn <- ifelse(!is.na(cm["0","1"]), cm["0","1"], 0)
  
  precision <- ifelse((tp + fp) > 0, tp / (tp + fp), 0)
  recall    <- ifelse((tp + fn) > 0, tp / (tp + fn), 0)
  f1        <- ifelse((precision + recall) > 0,
                      2 * precision * recall / (precision + recall), 0)
  accuracy  <- (tp + tn) / (tp + fp + tn + fn)
  
  data.frame(
    Threshold = t,
    Accuracy  = round(accuracy, 4),
    Precision = round(precision, 4),
    Recall    = round(recall, 4),
    F1_Score  = round(f1, 4)
  )
}))

kable(threshold_results, caption = "Model Performance Across Decision Thresholds")
```

**Business interpretation:**
- A **lower threshold (0.3)** prioritizes catching fraud — useful when the cost of a missed fraud is high (e.g. large transactions)
- A **higher threshold (0.7)** reduces false positives — useful when customer experience matters more and false declines are costly
- For most fraud systems, **0.4–0.5** balances both concerns well

---

# 8. Model Comparison & Conclusion

## 8.1 Model Comparison

```{r model-comparison}
# Consolidate all model metrics in one place for direct comparison.
# This gives a clear view of the tradeoffs between interpretability and performance.

# Logistic Regression metrics (from test set)
lr_preds <- ifelse(test_probs > 0.5, 1, 0)
lr_cm    <- table(Predicted = lr_preds, Actual = test_data$fraud)
lr_acc   <- sum(diag(lr_cm)) / sum(lr_cm)
lr_prec  <- lr_cm["1","1"] / sum(lr_cm["1",])
lr_rec   <- lr_cm["1","1"] / sum(lr_cm[,"1"])
lr_f1    <- 2 * lr_prec * lr_rec / (lr_prec + lr_rec)
lr_auc   <- as.numeric(auc(roc(test_data$fraud, test_probs)))

# Random Forest metrics (from CV model test set)
rf_cm_obj <- confusionMatrix(cv_test_preds, test_data_rf$fraud)
rf_acc    <- rf_cm_obj$overall["Accuracy"]
rf_prec   <- rf_cm_obj$byClass["Precision"]
rf_rec    <- rf_cm_obj$byClass["Recall"]
rf_f1     <- rf_cm_obj$byClass["F1"]

comparison_table <- data.frame(
  Model = c("Logistic Regression", "Random Forest (CV)"),
  Accuracy  = round(c(lr_acc, rf_acc), 4),
  Precision = round(c(lr_prec, rf_prec), 4),
  Recall    = round(c(lr_rec, rf_rec), 4),
  F1_Score  = round(c(lr_f1, rf_f1), 4),
  AUC       = c(round(lr_auc, 4), "N/A"),
  Interpretable = c("Yes", "No")
)

kable(comparison_table, caption = "Model Comparison: Logistic Regression vs Random Forest")
```

## 8.2 Recommendation

```{r recommendation, echo=FALSE}
cat("
RECOMMENDATION
==============
Both models perform well, but serve different purposes:

Logistic Regression:
- Fully interpretable — coefficients directly explain why a transaction is flagged
- Suitable for regulated environments where explainability is required
- AUC confirms strong discrimination ability
- Recommended for: production fraud flagging with human review

Random Forest:
- Higher raw accuracy, confirmed stable via 3-fold cross-validation
- Captures non-linear patterns and interactions automatically
- Less interpretable — harder to explain to stakeholders why a transaction was flagged
- Recommended for: automated batch scoring where throughput matters more than explainability

Final recommendation: deploy logistic regression as the primary model for
explainability, with RF as a secondary validation layer for high-value transactions.
")
```

## 8.3 Limitations & Next Steps

**Current limitations:**
- Dataset is synthetic — real-world fraud patterns are more complex and evolving
- Bootstrapping is a simple balancing approach; SMOTE may produce better minority class representation
- Models are static — fraud patterns shift over time, requiring periodic retraining

**What I would do next:**
1. **SMOTE** instead of bootstrapping for more realistic minority class augmentation
2. **XGBoost** — often outperforms RF on tabular fraud data with less tuning
3. **Real-time scoring pipeline** — wrap the model in a REST API (Plumber in R or FastAPI in Python) for live transaction scoring
4. **Drift monitoring** — track model performance over time as fraud patterns evolve
5. **Cost-sensitive learning** — assign different misclassification costs to false negatives vs false positives based on transaction value
